---
title: "Fitness_analysis"
author: "Cecilia Trivellin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of the genetic markers of microbial robustness

This script is used to analyse fitness data of Saccharomyces cerevisiae strains. Each strain carries a deletion that should confer either lower or higher robusntess. The genes that have been deleted are the following: 

1)	CEN.PK113-7D (ΔQDR1)
2)	CEN.PK113-7D (ΔGBP2)
3)	CEN.PK113-7D (ΔMRP13)
4)	CEN.PK113-7D (ΔWWM)
5)	CEN.PK113-7D (ΔHCM1)
6)	CEN.PK113-7D (ΔOCA4)
7)	CEN.PK113-7D (ΔRPS14A) 
8)	CEN.PK113-7D (ΔSMA2) 
9)	CEN.PK113-7D (ΔMSH3) 
10)	CEN.PK113-7D (ΔRPS14B) 
11)	CEN.PK113-7D (ΔMET28) 
12)	CEN.PK113-7D (ΔBCH1)
13)	CEN.PK113-7D (ΔTIR3) 
14)	CEN.PK113-7D (ΔHLJ1) 

The strains above were grown in three different perturbation spaces. THe first one containing the following perturbations: 

Delft
Delft_80NaCl
Delft_50etoh
Delft_25NaCl
Delft-05VM
Delft_100glu
Dleft_100etoh
Delft_5glu
Delft_NB
Delft_10LA
Delft_HMF1
Delft_FA2
Delft_pH3
Delft_AA2
SCH100
eSSH20

## Importing data and summary 
```{r}
library(readxl)
summary <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/media_composition.xlsx", 
    sheet = "summary")

QDR1_GBP2 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/LHFEPP_part1.xlsx", 
    sheet = "QDR1-GBP2_GV")

MRP13_WWM1 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/LHFEPP_part1.xlsx", 
    sheet = "MRP13-WWM1_GV")


HCM1_PCA4 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/LHFEPP_part1.xlsx", 
    sheet = "MRP13-WWM1_GV")


```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
