---
title: "Fitness_analysis"
author: "Cecilia Trivellin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of the genetic markers of microbial robustness

This script is used to analyse fitness data of Saccharomyces cerevisiae strains. Each strain carries a deletion that should confer either lower or higher robusntess. The genes that have been deleted are the following: 

1)	CEN.PK113-7D (ΔQDR1)
2)	CEN.PK113-7D (ΔGBP2)
3)	CEN.PK113-7D (ΔMRP13)
4)	CEN.PK113-7D (ΔWWM)
5)	CEN.PK113-7D (ΔHCM1)
6)	CEN.PK113-7D (ΔOCA4)
7)	CEN.PK113-7D (ΔRPS14A) 
8)	CEN.PK113-7D (ΔSMA2) 
9)	CEN.PK113-7D (ΔMSH3) 
10)	CEN.PK113-7D (ΔRPS14B) 
11)	CEN.PK113-7D (ΔMET28) 
12)	CEN.PK113-7D (ΔBCH1)
13)	CEN.PK113-7D (ΔTIR3) 
14)	CEN.PK113-7D (ΔHLJ1) 
15) CEN.PK113-7D parental
16) Ethanol Red

The strains above were grown in three different perturbation spaces. THe first one containing the following perturbations: 

Delft
Delft_80NaCl
Delft_50etoh
Delft_25NaCl
Delft-05VM
Delft_100glu
Dleft_100etoh
Delft_5glu
Delft_NB
Delft_10LA
Delft_HMF1
Delft_FA2
Delft_pH3
Delft_AA2
SCH100
eSSH20

### Loading libraries
```{r}
library(readxl)
library(tidyverse)
library(deSolve)
library(growthrates)
library(bbplot)
library(ggpubr)
```

## Importing data and summary 
```{r}
summary <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/media_composition_LHFEPP.xlsx", 
    sheet = "summary") 

strain <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/media_composition_LHFEPP.xlsx", 
    sheet = "strain") 


QDR1_GBP2 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part1.xlsx", 
    sheet = "QDR1-GBP2_GV") %>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 

MRP13_WWM1 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part1.xlsx", 
    sheet = "MRP13-WWM1_GV")%>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 


HCM1_OCA4 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part1.xlsx", 
    sheet = "HCM1-OCA4_GV")%>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 

RPS14A_SMA2 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part2.xlsx", 
    sheet = "RPS14A-SMA2_GV")%>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 

MSH3_RPS14B <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part2.xlsx", 
    sheet = "MSH3-RPS14B_GV")%>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 

BCH1_MET28 <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part2.xlsx", 
    sheet = "BCH1-MET28_GV")%>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 

HLJ1_ETRED <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part3.xlsx", 
    sheet = "HLJ1-ETRED_GV")%>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 

TIR3_CENPK <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/Fitness assays/prj_rob_genes_LHPS/LHFEPP_part3.xlsx", 
    sheet = "TIR3-CENPK_GV") %>%
    mutate(timeh = `time`/60) %>%
    select(-`time`)%>%
    gather("well", "value", -timeh) 
```
## Preprocessing growth data 
```{r}
ss_df <- function(df1, df2, df3, coln) {
  tmp <- inner_join(df1, df2, by="well")
  merge(tmp, df3[, c("well", coln)], by="well",all.x=TRUE)%>%
  group_by(media, timeh) %>%
  mutate(log_value=log(value)) %>%
  mutate(mean=mean(log_value)) %>%
  mutate(stdev=sd(log_value)) 
  # filter(replicate == 1) 
}
QDR1_GBP2 <- ss_df(QDR1_GBP2, summary, strain, "plate1") %>%
  rename(strain=plate1)
MRP13_WWM1 <- ss_df(MRP13_WWM1, summary, strain, "plate2")%>%
  rename(strain=plate2)
HCM1_OCA4 <- ss_df(HCM1_OCA4, summary, strain, "plate3")%>%
  rename(strain=plate3)
RPS14A_SMA2 <- ss_df(RPS14A_SMA2, summary, strain, "plate4")%>%
  rename(strain=plate4)
MSH3_RPS14B <- ss_df(MSH3_RPS14B, summary, strain, "plate5")%>%
  rename(strain=plate5)
BCH1_MET28 <- ss_df(BCH1_MET28, summary, strain, "plate6")%>%
  rename(strain=plate6)
HLJ1_ETRED <- ss_df(HLJ1_ETRED, summary, strain, "plate7")%>%
  rename(strain=plate7)
TIR3_CENPK <- ss_df(TIR3_CENPK, summary, strain, "plate8")%>%
  rename(strain=plate8)

df <- QDR1_GBP2 %>%
  rbind(MRP13_WWM1) %>%
  rbind(HCM1_OCA4) %>%
  rbind(RPS14A_SMA2)%>%
  rbind(MSH3_RPS14B)%>%
  rbind(BCH1_MET28) %>%
  rbind(HLJ1_ETRED) %>%
  rbind(TIR3_CENPK) 

saveRDS(df, file = "LHPS.rds")
```
## Checking growth curves with plots
```{r}
df <- df %>%
  mutate(replicate= as.character(replicate)) 
ggplot() +
  geom_point(data = subset(df, strain == c("HLJ1", "ETRED","TIR3", "CENPK")), size = 0.7, aes(y=log_value, x=timeh, colour=strain, shape=replicate)) +
scale_colour_brewer(palette = "Spectral") +
  labs(y = "log_GV", x = "time(h)")+
  facet_wrap(~media)  +
  bbc_style()

# ggsave("part3_growtcurves_replicates.pdf", width = 15, height = 10, path = "~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/R_scripts/genetic_markers_robustness/plots/LHPS_fitness")
```
## calculation of growth rate and lag phase 
```{r}
many_spline_fits <- all_splines(value ~ timeh | strain + media + replicate,
                                data = df, spar = 0.7)
par(mfrow = c(6,6))
par(mar=c(1,1,1,1))
plot(many_spline_fits)
#data frame is created with the results coming from all_splines function
results <- results(many_spline_fits) %>%
 mutate(mumax = ifelse(r2 <= 0.98, NA, mumax))%>%
  select(-`y0`)

results <- results %>%
  mutate(replicate= as.character(replicate)) %>%
  mutate(strain = factor(strain,levels=c("QDR1","MET28","MRP13","HCM1","GBP2","RPS14A","RPS14B","OCA4","MSH3", "BCH1", "WWM1","HLJ1", "TIR3", "SMA2", "CENPK", "ETRED"))) 

 p <- ggplot(data= results, 
      aes(x = strain,
          y = mumax, fill = strain)
    ) +
    ylim(0,0.3)+
    geom_violin(width=0.9,alpha = 0.3) +
    geom_boxplot(width=0.1,position=position_dodge(0.9), outlier.shape=NA, aes(middle=mean(mumax)))+
    scale_fill_manual(values=c("#922A2B","#922A2B","#922A2B","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140", "#4775A5", "#4775A5", "#4775A5", "#4775A5", "#4775A5","#5B9589","#5B9589")) +
    # facet_grid(cols =vars(Strain), rows = vars(Experiment)) +
    bbc_style()+
  stat_compare_means(method = "wilcox.test", label = "p.signif", 
                     ref.group = "CENPK", size=8)+
  guides(fill="none")+
    theme(strip.text = element_text(face="bold"),
        text = element_text(size = 15),
        axis.text.x = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1))

#ggsave("mumax_distribution_rawdata.pdf", width = 15, height = 10, path = "~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/R_scripts/genetic_markers_robustness/plots/LHPS_fitness")
```

## robusntess calculations 
```{r}
df_mean <- results %>%
  group_by(strain) %>% #calculating mean for each sample in the agar plate 
  mutate(mean=mean(mumax,  na.rm=TRUE)) %>%
  mutate(stdev=sd(mumax,  na.rm=TRUE)) %>%
  filter(media == "Delft")
mean_mumax <- mean(df_mean$mean)

df_R <- df_mean %>%
  mutate(robustness=-stdev^2/mean, na.rm=TRUE)%>%
  mutate(robustness_n=robustness/mean_mumax, na.rm=TRUE) %>%#calculating normalized robustness with the mean of all strains and all experiments
  group_by(strain) %>%
  mutate(mean_R=mean(robustness_n, na.rm=TRUE)) %>% #mean of the 4 replicates per sample coming from the 4 different plates
  mutate(sem_R=sd(robustness_n)/sqrt(3),na.rm=TRUE) %>% #standard error = standard deviation of the sampling distribution/sqrt n of samples
  filter(replicate == "1")
```
## plotting robusntess 
```{r}
p + geom_point(data=df_R, aes(x=strain, y=robustness_n, color=strain),size=5)+
  ylim(-0.2,0.3)+
  scale_color_manual(values=c("#922A2B","#922A2B","#922A2B","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140", "#4775A5", "#4775A5", "#4775A5", "#4775A5", "#4775A5","#5B9589","#5B9589")) +
  geom_hline(aes(yintercept = robustness_n), 
              data = filter(df_R, strain == "CENPK"),
             linetype = "dashed", color = "#5B9589")+
  bbc_style()+
  guides(color="none")+
  theme(strip.text = element_text(face="bold"),
        text = element_text(size = 15),
        axis.text.x = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1))
ggsave("FR_LHPS.pdf", width = 15, height = 10, path = "~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/R_scripts/genetic_markers_robustness/plots/LHPS_fitness")
```

