---
title: "FR_overall_BPS+LHPS+CPS"
author: "Cecilia Trivellin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE} 
#Set up the R environment
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Analysis of fitness and robusntess of all three perturbation spaces 
(see other scripts and material and methods in publication) 

### Loading libraries
```{r}
library(readxl)
library(tidyverse)
library(deSolve)
library(lattice)
library(growthrates)
library(bbplot)
library(ggpubr)
library(Cairo)
```

### Importing data
```{r}
LHPS_df <- readRDS("~/your_directory/LHPS.rds") %>%
  mutate(p_space="LHPS")
BPS_df <- readRDS("~/your_directory/BPS.rds") %>%
  mutate(p_space="BPS")
CPS_df <- readRDS("~/your_directory/CPS.rds") %>%
  mutate(p_space="CPS")
```

### Merging the results
```{r}
df <- LHPS_df %>%
  rbind(BPS_df) %>%
  rbind(CPS_df)

head(df)
```

### Calculating max. specific growth rates
```{r}
df <- na.omit(df) #because I was getting an error of missing values insmooth splines
# Fit spline models for each combination of 'strain', 'media', and 'replicate'
many_spline_fits <- all_splines(value ~ timeh | strain + media + replicate + p_space, data = df, spar = 0.7)

results <- results(many_spline_fits) 
# par(mfrow = c(6,6))
# par(mar=c(1,1,1,1))
# plot(many_spline_fits)

#data frame is created with the results coming from all_splines function
#based on the single environments set to zero the strains that did not grow (as done in the single scripts)
#LHPS
results$mumax <- ifelse(results$media == "Delft_100etoh" | results$media == "Delft_pH3", 0 , results$mumax) #none of the strains grew with Delft_100etoh or pH3
#single strains that did not grow in certain conditions 
results$mumax <- ifelse((results$media == "Delft_80NaCl" | results$media == "Delft_FA2" | results$media == "SCH100") & results$strain == "met28Δ", 0 , results$mumax)
#BPS had none
#CPS 
#the umax of the strains that did not grow is set to 0 based on the growth curves above
results$mumax <- ifelse(results$media == "CPS_14" , 0 , results$mumax)
results$mumax <- ifelse(results$media == "CPS_2" & results$strain == "msh3Δ" & results$replicate == "3", 0 , results$mumax) 
results$mumax <- ifelse(results$media == "CPS_7" & results$strain == "wwm1Δ" , 0 , results$mumax) 
results$mumax <- ifelse(results$media == "CPS_5" & (results$strain == "rps14aΔ" | results$strain == "tir3Δ" | results$strain == "CEN.PK113-7D") & (results$replicate == "2" | results$replicate == "3"), 0 , results$mumax) 
#single strains that did not grow in certain conditions 
results$mumax <- ifelse((results$media == "Delft_25NaCl" | results$media == "Delft_FA2" | results$media == "SCH100") & results$strain == "met28Δ", 0 , results$mumax)

# now the mumax of remaining strains with r2 <= 0.99 and mumax != 0 is set to NA
 results <- results %>%
   mutate(mumax = ifelse(r2 <= 0.99 & mumax != 0, NA, mumax))%>% #first I convert all the mumax values <0.99 to NA as the approximation of the function is not reiable 
  select(-`y0`)
```
 
### Visualizing overall results
```{r}
results <- results %>%
  mutate(replicate= as.character(replicate)) %>%
  mutate(strain = factor(strain,levels=c("CEN.PK113-7D","EthanolRed","qdr1","met28","mrp13","hcm1","gbp2","rps14a","rps14b","oca4","msh3", "bch1", "wwm1","hlj1", "tir3", "sma2"))) 

head(results)

# Calculate the mean 'mumax' for the "CEN.PK113-7D" strain
mean_mumax_CENPK <- mean(filter(results, strain == "CEN.PK113-7D")$mumax)

p <- ggplot(data= results, 
      aes(x = strain,
          y = mumax, fill = strain)
    ) +
    ylim(0,0.3)+
    geom_violin(width=0.9,alpha = 0.3) +
    geom_boxplot(width=0.1,position=position_dodge(0.9), outlier.shape=NA)+
  stat_summary(fun=mean, geom='point', shape=20, size=3, color="white") +
    scale_fill_manual(values=c("black","black","#922A2B","#922A2B","#922A2B","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140", "#4775A5", "#4775A5", "#4775A5", "#4775A5", "#4775A5")) +
    # facet_grid(cols =vars(Strain), rows = vars(Experiment)) +
  geom_hline(yintercept = mean_mumax_CENPK, linetype = "dashed", color = "black")+
    bbc_style()+
  stat_compare_means(method = "wilcox.test", label = "p.signif", 
                     ref.group = "CEN.PK113-7D", size=8, hide.ns = TRUE)+
  guides(fill="none")+
    theme(strip.text = element_text(face="bold"),
        text = element_text(size = 15),
        axis.text.x = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1))
```

### Robusntess calculations (Trivellin et.al.,2022)
```{r}
df_mean <- results %>%
  group_by(strain) %>% #calculating mean for each sample in the agar plate 
  mutate(mean=mean(mumax,  na.rm=TRUE)) %>%
  mutate(stdev=sd(mumax,  na.rm=TRUE)) %>%
  filter(media == "BPS_1")
mean_mumax <- mean(df_mean$mean) 

df_R <- df_mean %>%
  mutate(robustness=-stdev^2/mean, na.rm=TRUE)%>%
  mutate(robustness_n=robustness/mean_mumax, na.rm=TRUE) %>%#calculating normalized robustness with the mean of all strains and all experiments
  group_by(strain) %>%
  mutate(mean_R=mean(robustness_n, na.rm=TRUE)) %>% #mean of the 4 replicates per sample coming from the 4 different plates
  mutate(sem_R=sd(robustness_n)/sqrt(3),na.rm=TRUE) %>% #standard error = standard deviation of the sampling distribution/sqrt n of samples
  filter(replicate == "1")

head(df_R)
```
### Plotting robusntess 
```{r}
p + geom_point(data=df_R, aes(x=strain, y=robustness_n, color=strain),size=5)+
  ylim(-0.35,0.3)+
  scale_color_manual(values=c("black","black","#922A2B","#922A2B","#922A2B","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140","#E2C140", "#4775A5", "#4775A5", "#4775A5", "#4775A5", "#4775A5")) +
  geom_hline(aes(yintercept = robustness_n), 
              data = filter(df_R, strain == "CEN.PK113-7D"),
             linetype = "dashed", color = "black")+
  bbc_style()+
  guides(color="none")+
  theme(strip.text = element_text(face="bold"),
        text = element_text(size = 15),
        axis.text.x = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1))
#ggsave("FR_overall.pdf", device= cairo_pdf, width = 15, height = 10, path = "~/your_directory/")
```
Checking mumax in all perturbations to see which strain stands out 
```{r}
# copnverting 'media' to factor to order perturbations
results <- results %>%
  mutate(media = factor(media,levels=c("Delft", "Delft_5glu", "Delft_100glu","Delft_NB", "Delft_pH3", "Delft-05VM", "Delft_25NaCl", "Delft_80NaCl",  "Delft_50etoh",  "Delft_100etoh", "Delft_AA2", "Delft_FA2",  "Delft_10LA", "Delft_HMF1", "eSSH20", "SCH100", "BPS_1", "BPS_2",  "BPS_3", "BPS_4", "BPS_5", "BPS_6", "BPS_7", "BPS_8", "BPS_9", "BPS_10", "BPS_11", "BPS_12", "BPS_13", "BPS_14", "BPS_15", "BPS_16", "CPS_1", "CPS_2",  "CPS_3", "CPS_4", "CPS_5", "CPS_6", "CPS_7", "CPS_8", "CPS_9", "CPS_10", "CPS_11", "CPS_12", "CPS_13", "CPS_14", "CPS_15", "CPS_16"))) %>%
  mutate(strain = factor(strain,levels=c("CEN.PK113-7D","EthanolRed","qdr1","met28","mrp13","hcm1","gbp2","rps14a","rps14b","oca4","msh3", "bch1", "wwm1","hlj1", "tir3", "sma2"))) 
      
ggplot(results, aes(x = media, y = strain)) + 
  geom_tile(colour="black", size = 0.25, aes(fill = mumax)) +
  scale_fill_gradientn(colours = c("#4775A5", "#7acfbd","#E9C46A", "#922A2B"),
                       na.value = "#C9C9C9") +
  coord_fixed(ratio = 1)+
  theme(axis.text = element_text(size = 10),
        axis.text.x = element_text(size = 10, angle=90, vjust = 0.5,hjust = 1))

#ggsave("fitness_heatmap.pdf", device=cairo_pdf, width = 15, height = 10, path = "~/your_directory/")
```
### Robusntess overall 
```{r}
results <- results %>%
  mutate(p_space = case_when(
    grepl("^CPS", media) ~ "CPS",
    grepl("^BPS", media) ~ "BPS",
    TRUE ~ "LHPS"
  ))

df_mean_2 <- results %>%
  group_by(strain, p_space) %>% #calculating mean for each sample 
  mutate(mean=mean(mumax,  na.rm=TRUE)) %>%
  mutate(stdev=sd(mumax,  na.rm=TRUE)) %>%
  filter((media == "CPS_1" | media == "BPS_1" | media == "Delft")& replicate==1)

# calculating mean for each perturbation space
mean_mumax_BPS <- mean(df_mean_2$mean[df_mean_2$p_space == "BPS"])
mean_mumax_CPS <- mean(df_mean_2$mean[df_mean_2$p_space == "CPS"])
mean_mumax_LHPS <- mean(df_mean_2$mean[df_mean_2$p_space == "LHPS"])

df_R_2 <- df_mean_2 %>%
  mutate(mean_p_space = case_when(
    p_space %in% "BPS" ~ mean_mumax_BPS,
    p_space %in% "CPS" ~ mean_mumax_CPS,
    p_space %in% "LHPS" ~ mean_mumax_LHPS)) %>%
  mutate(robustness=-stdev^2/mean, na.rm=TRUE)%>%
  mutate(robustness_n=robustness/mean_p_space, na.rm=TRUE) #calculating normalized robustness with the mean of all strains and all experiment

head(df_R_2)
```
Plotting as heatmap
```{r}
df_R_2 <- df_R_2 %>%
  mutate(strain = factor(strain,levels=c("CEN.PK113-7D","EthanolRed","qdr1","met28","mrp13","hcm1","gbp2","rps14a","rps14b","oca4","msh3", "bch1", "wwm1","hlj1", "tir3", "sma2"))) 
df_R <- df_R %>%
  mutate(p_space="overall") 
try <- bind_rows(df_R_2, df_R) %>%
  mutate(p_space = factor(p_space,levels=c("LHPS", "BPC", "CPS", "overall"))) 
      
ggplot(try, aes(x = p_space, y = strain)) + 
  geom_tile(colour="black", size = 0.25, aes(fill = robustness_n)) +
  scale_fill_gradientn(colours = c("#4775A5", "#7acfbd","#E9C46A", "#922A2B"),
                       na.value = "#C9C9C9") +
  coord_fixed(ratio = 1)+
  theme(axis.text = element_text(size = 20),
        axis.text.x = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1))

#ggsave("robustness_heatmap.pdf", device=cairo_pdf, width = 15, height = 10, path = "~/your_directory/")
```


