---
title: "Robusntess on yeast genetic interaction network"
author: "Cecilia Trivellin"
date: "`r Sys.Date()`"
output: html_document
---
### Industrial Biotechnology Division, Chalmers University of Technology 2023

## Aim of the script
The datasets used in this script were taken from Costanzo et.al. 2021 and used to calculate robustness for each mutant according to the formula in Trivellin et.al 2022.

### Load libraries
```{r, results='hide'}
library(tidyverse)
library(readxl)
library(bbplot)
library(ggplot2)
library(ggvenn)
library(VennDiagram)
library(ggpubr)
library(packcircles)
library(RColorBrewer)
library(patchwork)
library(treemapify)
```
### Load dataset
```{r}
# Read Excel file into a dataframe named 'df'
df <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/raw data/Costanzo_et_al_Data_File_1_Conditions_Strains_Fitness.xlsx", 
    sheet = "Mutant Fitness_Conditions") %>%
    # Rename the column "Strain ID" to "ID"
    rename("ID" = "Strain ID")

# Replace missing values in columns 6 to 19 with 0
df[, 6:19][is.na(df[, 6:19])] <- 0

# Separate the column "Allele (Essential genes only)" into two columns "gene" and "allele"
# Select all columns except the original "gene"
# Combine columns from "Gene Name" to "allele" into a new column "gene" separated by "-"
df <- df %>% 
  separate(`Allele (Essential genes only)`, c("gene", "allele"), "-") %>%
  select(-`gene`) %>%
  unite('gene', `Gene Name`:`allele`, sep = "-", remove = TRUE, na.rm = TRUE)
```

## Explore fitness across conditions 
```{r}
# Gather the dataframe 'df' from wide to long format using specific columns
df_long <- df %>%
  gather(key = "condition", value = "value", c("Actinomycin D", "Benomyl", "Boretzeomib", "Caspofungin", "Concanmycin A", "Cycloheximide", "Fluconazole", "Galactose", "Geldanamycin", "MMS", "Monensin", "Rapamycin", "Sorbitol", "Tunicamycin"))

# Create a violin plot and box plot using ggplot2
violin_box_plot <- ggplot(df_long, aes(x = condition, y = value)) + 
  geom_violin(trim = TRUE) +
  geom_boxplot(width = 0.1, outlier.shape = NA, aes(middle = mean(value))) +
  bbc_style() + 
  labs(title = "Fitness distribution", y = "normalized colony size", x = "growth condition") +
  stat_compare_means(method = "anova", label.y = 1.5) +  # Add ANOVA test comparison
  theme(legend.position = "none",
        axis.text.x = element_text(size = 20, angle = 90, vjust = 0.5, hjust = 1),
        text = element_text(size = 15),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 20, face = "bold"))

# Calculate overall distribution using density estimation
overall_density_dist <- density(df_long$value)

# Convert density data to a data frame
overall_density_df <- data.frame(x = overall_density_dist$x, y = overall_density_dist$y)

# Create a plot for the overall distribution
overall_distribution_plot <- ggplot(overall_density_df, aes(x = x, y = y)) +
  geom_line(color = "black", size = 1.2) +
  bbc_style() +  
  labs(title = "Overall Distribution", x = "normalized colony size", y = "Density") +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 20, angle = 90, vjust = 0.5, hjust = 1),
        text = element_text(size = 15),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 20, face = "bold"))

# Combine the two plots using patchwork
combined_plot <- violin_box_plot + overall_distribution_plot +
  plot_layout(ncol = 2, widths = c(3, 1))  # Adjust the widths as per your preference

# Display the combined plot
print(combined_plot)

# Save the combined plot as a PDF file
# ggsave("fitness_distribution_overall.pdf", width = 17, height = 10, path = "~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/R_scripts/genetic_markers_robustness/plots/analysis_costanzo")

```
## Calculating Robustness
Calculating mean and stdev for each strain 
```{r}
df_1 <- df %>% 
  mutate(mean = rowMeans(select(df, 5:18), na.rm = TRUE))
df_1$sd = apply(df_1[,5:18], 1, sd, na.rm=TRUE)
```
Computing the Fano factor for each strain
```{r}
df_R <- df_1 %>%
  mutate(R=-sd^2/mean)
```
Normalizing the Fano factor for the mean value of the fitness across all strains 
```{r}
meanF <- mean(df_R$mean)

df_R_norm <- df_R %>%
  mutate(R_n=R/meanF)
```
Plotting robustness with and without normalization with fitness mean across all samples
```{r}
df_long_R <- df_R_norm %>% 
  gather(key="robustness", value="value", c("R", "R_n")) 
# checking data distribution 
ggplot(df_long_R, aes(x = robustness, y = value)) + 
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 20, angle=90, vjust = 0.5,hjust = 1),
        text = element_text(size = 20), axis.text=element_text(size=20),
        axis.title=element_text(size=20,face="bold"))
#ggsave("RvsRn.png", width = 500, height = 300, unit = "mm", dpi = 300)

```
### Extracting top and last 10 strains based on the normalized robustness
```{r}
# Selecting the last 10 strains based on normalized Fano factor
df_R_last_10 <- df_R_norm %>% 
    arrange(R_n) %>%  # Arrange in ascending order based on R_n
    slice(1:10)  # Select the last 10 rows

# Selecting strains with R_n equal to 0
df_R_zero <- df_R_norm[df_R_norm$R_n == 0,]  # Filter strains with R_n = 0

# Selecting the top 10 strains with R_n not equal to 0, ordered by descending R_n
df_R_top_10 <- filter(df_R_norm, R_n < 0) %>%
  arrange(desc(R_n)) %>%
  slice(1:10)  # Select the top 10 strains

# Selecting the best and worst 10 strains based on Fitness (mean)
df_F_high_10 <- filter(df_R_norm) %>%
  arrange(desc(mean)) %>%
  slice(1:10)  # Select the top 10 strains based on mean

df_F_low_10 <- filter(df_R_norm) %>%
  arrange(mean) %>%
  slice(1:10)  # Select the bottom 10 strains based on mean

```
### Venn diagram with 10th and 90th percentiles
```{r}
# Isolating the 10th and 90th percentiles with best/worst fitness and best/worst robustness
best_90_F <- df_R_norm %>%
  filter(mean >= quantile(df_R_norm$mean, 0.90))  # Select strains with fitness in the top 10%

best_90_R <- df_R_norm %>%
  filter(R_n >= quantile(df_R_norm$R_n, 0.90))  # Select strains with robustness in the top 10%

worst_10_F <- df_R_norm %>%
  filter(mean <= quantile(df_R_norm$mean, 0.10))  # Select strains with fitness in the bottom 10%

worst_10_R <- df_R_norm %>%
  filter(R_n <= quantile(df_R_norm$R_n, 0.10))  # Select strains with robustness in the bottom 10%

# Create a Venn diagram to visualize gene overlap between different categories
venn <- list(High_Fitness = best_90_F$ID, High_Robustness = best_90_R$ID, Low_Fitness = worst_10_F$ID, Low_Robustness = worst_10_R$ID)
ggvenn(venn, fill_color = c("#922A2B", "#E2C140", "#4775A5", "#5B9589"), text_size = 8, set_name_size = 10)

# Save the Venn diagram as a PDF file (optional)
# ggsave("venn_gene_overlap_quantile.pdf", width = 20, height = 10, path = "~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/R_scripts/genetic_markers_robustness/plots")

# Calculate the overlap between different sections of the Venn diagram
overlap <- calculate.overlap(x = venn)

# Extracting strains from the sections of interest
hFhR <- df_R_norm[df_R_norm$ID %in% overlap[[6]],]  # High Fitness and High Robustness
lFlR <- df_R_norm[df_R_norm$ID %in% overlap[[11]],]  # Low Fitness and Low Robustness
hFlR <- df_R_norm[df_R_norm$ID %in% overlap[[8]],]   # High Fitness and Low Robustness
hRlF <- df_R_norm[df_R_norm$ID %in% overlap[[9]],]   # High Robustness and Low Fitness

```
### Exploring correlations between robusntess and performance 
```{r}
# Create a scatter plot with a regression line, confidence interval, and correlation coefficient
# using ggscatter from the ggpubr package
ggscatter(df_R_norm, x = "R_n", y = "mean",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Robustness_normalized", ylab = "Fitness (colony size)")

# Cut the data between -0.2 and 0 for improved visualization
df_Rcorr <- filter(df_R_norm, R_n > -0.1) 

# Create another scatter plot with a regression line, confidence interval, and correlation coefficient
# for the subset of data where R_n is greater than -0.1
ggscatter(df_Rcorr, x = "R_n", y = "mean", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Robustness_normalized", ylab = "Fitness (colony size)")

```
### Plotting robusntess and fitness to compare with experimental studies
```{r}
# Extract strains of interest from the fitness dataset
filtered_strain <- c("qdr1", "met28", "mrp13", "hcm1", "gbp2", "rps14a", "rps14b", "oca4", "msh3", "bch1", "wwm1", "hlj1", "tir3", "sma2")
fitness_strainsoi <- df_long %>%
  filter(gene %in% filtered_strain) 

# Ordering strains for plotting
fitness_strainsoi <- fitness_strainsoi %>%
  mutate(gene = factor(gene, levels = c("qdr1", "met28", "mrp13", "hcm1", "gbp2", "rps14a", "rps14b", "oca4", "msh3", "bch1", "wwm1", "hlj1", "tir3", "sma2")))

p <- ggplot(data = fitness_strainsoi, 
      aes(x = gene,
          y = value, color = gene)
    ) +
    geom_point(size = 5) +
    stat_summary(fun = mean, geom = 'point', shape = 20, size = 3, color = "black") +
    scale_colour_manual(values = c("#922A2B", "#922A2B", "#922A2B", "#E2C140", "#E2C140", "#E2C140", "#E2C140", "#E2C140", "#E2C140", "#4775A5", "#4775A5", "#4775A5", "#4775A5", "#4775A5")) +
    bbc_style() +
    guides(fill = "none") +
    theme(strip.text = element_text(face = "bold"),
          text = element_text(size = 15),
          axis.text.x = element_text(size = 20, angle = 90, vjust = 0.5, hjust = 1))

# Extract strains of interest from the robustness dataset
rob_strainsoi <- df_long_R %>%
  filter(gene %in% filtered_strain) %>%
  filter(robustness == "R_n")

#ordering strains
rob_strainsoi <- rob_strainsoi %>%
  mutate(gene = factor(gene, levels = c("qdr1", "met28", "mrp13", "hcm1", "gbp2", "rps14a", "rps14b", "oca4", "msh3", "bch1", "wwm1", "hlj1", "tir3", "sma2")))

custom_breaks <- c(-0.8, -0.6, -0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8, 1, 1.2)

# Combine fitness and robustness plots
p + geom_point(data = rob_strainsoi, aes(x = gene, y = value, color = gene), size = 5, shape = 17) +
  scale_color_manual(values = c("#922A2B", "#922A2B", "#922A2B", "#E2C140", "#E2C140", "#E2C140", "#E2C140", "#E2C140", "#E2C140", "#4775A5", "#4775A5", "#4775A5", "#4775A5", "#4775A5")) +
  bbc_style() +
  scale_y_continuous(breaks = custom_breaks) +
  guides(color = "none") +
  theme(strip.text = element_text(face = "bold"),
        text = element_text(size = 15),
        axis.text.x = element_text(size = 20, angle = 90, vjust = 0.5, hjust = 1))

#ggsave("FR_referencedataset.pdf", device= cairo_pdf, width = 15, height = 10, path = "~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/R_scripts/genetic_markers_robustness/plots/overall_new")

```

### Exploring SAFE network regions 
```{r}
#loading dataset from Costanzo et.al 2021
SAFE <- read_excel("~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/raw data/Costanzo_et_al_Data_File_2_SMF_Functional_enrichment.xlsx", 
    sheet = "SAFE")
```
Comparing the top robust strains with SAFE 
```{r}
# Find the intersection of genes between the best 90 R strains and the SAFE dataset
topR_SAFE <- generics::intersect(best_90_R$gene, SAFE$`Gene/Allele`) 

# Extract rows corresponding to the top genes from the SAFE dataset
top_R_strains <- SAFE[SAFE$`Gene/Allele` %in% topR_SAFE,]

# Working on a separate dataset for R0 and merging
topR0_SAFE <- generics::intersect(df_R_zero$`gene`, SAFE$`Gene/Allele`)

# Extract rows corresponding to the top R0 genes from the SAFE dataset
top_R0_strains <- SAFE[SAFE$`Gene/Allele` %in% topR0_SAFE,]

# Combine the datasets for the top R strains and top R0 strains, and rename the SAFE network region column
top_R_strains <- top_R_strains %>%
  rbind(top_R0_strains) %>%
  rename(SAFE_group = `SAFE network region`)

# Count the occurrences of each SAFE network region for the top R strains
count_high_R_process <- top_R_strains %>%
  group_by(SAFE_group) %>%
  tally()

```
Comparing the worst robut strains with SAFE 
```{r}
# Find the intersection of genes between the worst 10 R strains and the SAFE dataset
worstR_SAFE <- generics::intersect(worst_10_R$`gene`, SAFE$`Gene/Allele`)

# Extract rows corresponding to the worst genes from the SAFE dataset
worst_R_strains <- SAFE[SAFE$`Gene/Allele` %in% worstR_SAFE,] %>%
  rename(SAFE_group = `SAFE network region`)

# Count the occurrences of each SAFE network region for the worst R strains
count_low_R_process <- worst_R_strains %>%
  group_by(SAFE_group) %>%
  tally()

```
Comparing the top fit strains with SAFE 
```{r}
# Find the intersection of genes between the best 90 F strains and the SAFE dataset
topF_SAFE <- generics::intersect(best_90_F$gene, SAFE$`Gene/Allele`) 

# Extract rows corresponding to the top genes from the SAFE dataset
top_F_strains <- SAFE[SAFE$`Gene/Allele` %in% topF_SAFE,] %>%
  rename(SAFE_group = `SAFE network region`)

# Count the occurrences of each SAFE network region for the top F strains
count_high_F_process <- top_F_strains %>%
  group_by(SAFE_group) %>%
  tally() 
```
Comparing the worst robust strains with SAFE 
```{r}
# Find the intersection of genes between the worst 10 F strains and the SAFE dataset
worstF_SAFE <- generics::intersect(worst_10_F$`gene`, SAFE$`Gene/Allele`)

# Extract rows corresponding to the worst genes from the SAFE dataset
worst_F_strains <- SAFE[SAFE$`Gene/Allele` %in% worstF_SAFE,] %>%
  rename(SAFE_group = `SAFE network region`)

# Count the occurrences of each SAFE network region for the worst F strains
count_low_F_process <- worst_F_strains %>%
  group_by(SAFE_group) %>%
  tally() 
```

### putting the four bubble plots together
```{r}
# Function to add an ID column and select the top 20 most represented SAFE regions
add_ID <- function(df, ID) {
  df %>%
    arrange(desc(n)) %>%
    slice(1:20) %>%
    mutate(ID = ID)
}

# Extracting overly represented processes for different categories
HR <- add_ID(count_high_R_process, "High Robustness")
LR <- add_ID(count_low_F_process, "Low Robustness")
HF <- add_ID(count_high_F_process, "High Fitness")
LF <- add_ID(count_low_F_process, "Low Fitness")

# Combine the data for plotting
df_bubble <- HR %>%
  rbind(LR) %>%
  rbind(HF) %>%
  rbind(LF)

# Perform treemap layout
packing_tot <- circleProgressiveLayout(df_bubble$n, sizetype = 'area')
packing_tot$radius <- 0.95 * packing_tot$radius
data_tot <- cbind(df_bubble, packing_tot)
dat.gg_tot <- circleLayoutVertices(packing_tot, npoints = 50) 

# Plot the treemap
ggplot(data_tot, aes(area = n, fill = ID, label = SAFE_group, subgroup = ID)) +
  geom_treemap(colour = "white", size = 2) +
  geom_treemap_subgroup_border(colour = "white", size = 5) +
  geom_treemap_text(grow = T, reflow = T, colour = "white") +
  facet_wrap( ~ ID) +
  scale_fill_manual(values = c("#922A2B", "#E2C140", "#4775A5", "#5B9589")) +
  theme(legend.position = "bottom") +
  labs(fill = "Group") +
  bbc_style()

# Save the treemap plot
#ggsave("treemap_overall.pdf", width = 230, height = 175, unit = "mm", path = "~/Library/CloudStorage/OneDrive-Chalmers/PhD/Experimental_Data/Robusntess application on existing dataset/R_scripts/genetic_markers_robustness/plots")

```


